---
title: "Homework 1"
author: "Balpreet Kaur Saluja - 24202918"
format: 
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    smaller: true
    css: custom.css
execute: 
  echo: true
engine: knitr
editor: visual
---

## Introduction:

We analyze social and development indicators from three diverse countries: **Brazil** (a large upper-middle-income country in South America), **Kazakhstan** (a Central Asian upper-middle-income nation), and **Madagascar** (a low-income island country in Africa). These countries were chosen to highlight contrasts in population trends, poverty, and life expectancy across different regions and income levels.

```{r echo=FALSE}
library(data.table)
library(ggplot2)
```

## 1. Read and preprocess the data:

```{r}
bra <- fread("indicators_bra.csv")
kaz <- fread("indicators_kaz.csv")
mdg <- fread("indicators_mdg.csv")
bra[, lapply(.SD, class)]
kaz[, lapply(.SD, class)]
mdg[, lapply(.SD, class)]
```

The first row in each dataset contains headers or empty info, so we remove it. We also ensure`Year` is integer and `Value` is numeric for valid analysis.

```{r}
bra <- bra[-1]
bra[, Year := as.integer(Year)]
bra[, Value := as.numeric(Value)]
kaz <- kaz[-1]
kaz[, Year := as.integer(Year)]
kaz[, Value := as.numeric(Value)]
mdg <- mdg[-1]
mdg[, Year := as.integer(Year)]
mdg[, Value := as.numeric(Value)]
```

## 2. Merging the data:

```{r}
# Add a Country column to each data.table (if not already present)
bra[, Country := "Brazil"]
kaz[, Country := "Kazakhstan"]
mdg[, Country := "Madagascar"]

# Merge the datasets using data.table's rbindlist()
merged_data <- rbindlist(list(bra, kaz, mdg))



```

A `Country` column is added to each dataset to keep track of origin. We merge the three data.tables using `rbindlist()` for fast vertical stacking.

## 3. Quick data exploration to know more about data:

```{r}
# Number of unique indicators in each country
merged_data[, uniqueN(`Indicator Name`), by = Country]
```

This shows the breadth of data coverage per country by counting how many unique indicators are available in each dataset.

```{r}
# Earliest and latest year available per country
merged_data[, .(MinYear = min(Year, na.rm = TRUE),
                MaxYear = max(Year, na.rm = TRUE)), 
            by = Country]
```

We explore the time span of data available for each country, helping us understand historical depth and recentness.

## 

```{r}
# Top 5 most frequent indicators overall
merged_data[, .N, by = `Indicator Name`][order(-N)][1:5]
```

Identifies which indicators are most frequently recorded across all countries — potentially useful for cross-country comparisons.

```{r}
# Summary statistics for Value by country
merged_data[, .(
  Min = min(Value, na.rm = TRUE),
  Q1  = quantile(Value, 0.25, na.rm = TRUE),
  Median = median(Value, na.rm = TRUE),
  Q3 = quantile(Value, 0.75, na.rm = TRUE),
  Max = max(Value, na.rm = TRUE)
), by = Country]

```

Basic summary statistics give an overview of the range and central tendency of values recorded per country.

## 4. Data Analysis Using `data.table (keyby):`

```{r}
# Selecting 3 indicators for focused analysis
focus_indicators <- c(
  "Population, total",
  "Life expectancy at birth, total (years)",
  "Poverty headcount ratio at $5.50 a day (2011 PPP) (% of population)"
)

focus_data <- merged_data[`Indicator Name` %in% focus_indicators]

#Average Value per Country & Indicator
focus_data[, .(
  Average = mean(Value, na.rm = TRUE),
  Count = .N
), keyby = .(`Indicator Name`, Country)]
```

We focus the analysis on three crucial indicators across all countries — population size, life expectancy, and poverty — for meaningful comparison.

## 

```{r}
focus_data[, .(
  MinYear = min(Year, na.rm = TRUE),
  MaxYear = max(Year, na.rm = TRUE)
), keyby = .(`Indicator Name`, Country)]

```

Here we check how long each indicator has been tracked in each country — useful for understanding data continuity.

```{r}
focus_data[is.na(Value), .N, keyby = .(`Indicator Name`, Country)]

```

This identifies missing values per indicator and country, helping assess data quality.

## 

**5. Plotting Results:**

```{r}
# Filter population data
pop <- focus_data[`Indicator Name` == "Population, total"]

ggplot(pop, aes(x = Year, y = Value / 1e6, color = Country)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Population Growth (in millions)", 
       y = "Population (millions)", x = "Year") +
  theme_minimal()


```

**Interpretation:** Brazil and Madagascar show consistent growth, with Madagascar having a steeper rise. Kazakhstan's growth is slower but steady.

## 

```{r}
# Preparing life data
life <- focus_data[`Indicator Name` == "Life expectancy at birth, total (years)"]

ggplot(life, aes(x = Country, y = Value, fill = Country)) +
  geom_boxplot() +
  labs(title = "Distribution of Life Expectancy by Country",
       y = "Years", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Interpretation:** Brazil has the highest median life expectancy, followed by Kazakhstan. Madagascar lags behind, reflecting differences in development.
